---
title: "Flight Delay""
author: "Duc Tran and William Wiik"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: yes
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Flight Delay"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
library(bonuslabpackage)
library(dplyr)
library(caret)
library(nycflights13)
```

# Introduction
This vignette shows the prediction of the function `ridgereg()` from the package `bonuslabpackage` on flight data from the package `nycflights13`.

# 1.2.1 Predictive modeling of flight delays using `ridgereg()`

We read in the weather dataset and the flights dataset from the `nycflights13` package and select
variables we believe to have a predictive value.

```{r}

 
flights <- nycflights13::flights
weather <- nycflights13::weather

# Nice variables
flights <- flights %>%  select(origin, year, month, day, hour,
                               dep_delay, arr_delay, distance)

weather <- weather %>%  select(origin, year, month,
                               day, hour, temp, wind_speed)
  
```


## 2

We add extra weather data (temp and wind_speed) from the weather dataset and create multiplicative interaction effects of temp and wind_speed.

We also make the assumption that the response variable is arrival delay.


```{r}

# merge flights with weather
merged_data <- flights %>%
  left_join(weather, by = c("origin", "year", "month", "day", "hour")) %>% 
  mutate(interactive_temp_wind = temp * wind_speed) %>% 
  na.omit() # remove Na's


```


## 3

We use the `caret` package to divide the flight dataset into three sets: test, train and validation (with the proportions 5%, 80% and 15%).


```{r}

# Set the seed for reproducibility
set.seed(123)

train_indices <- createDataPartition(merged_data$arr_delay, p = 0.80, list = FALSE)
train_set <- merged_data[train_indices, ]

test_val_set <- merged_data[-train_indices, ]


# 5% of 20% is 25%, thats why p = 0.25
test_indices <- createDataPartition(test_val_set$arr_delay, p = 0.25, list = FALSE)

test_set <- test_val_set[test_indices, ]
validation_set <- test_val_set[-test_indices, ]


```



## 4

We train ridge regressions models for different values of λ and evaluate the root mean squared error.
The formula for the model is: "arr_delay ~ dep_delay + distance + temp + wind_speed + interactive_temp_wind".

```{r}
index <-c(6,8,9,10,11)
predictors_subset <- colnames(merged_data)[index]
formula <- as.formula(paste("arr_delay ~", paste(predictors_subset, collapse = "+")))

ridge_1 <- ridgereg(formula, train_set, lambda=1)
ridge_5 <- ridgereg(formula, train_set, lambda=5)
ridge_10 <- ridgereg(formula, train_set, lambda=10)
ridge_50 <- ridgereg(formula, train_set, lambda=50)
ridge_200 <- ridgereg(formula, train_set, lambda=200)
ridge_10000 <- ridgereg(formula, train_set, lambda=10000)


resid_1 <- validation_set$arr_delay - ridge_1$predict(validation_set[index])
RMSE_1 <- sqrt(mean(resid_1^2)) 

resid_5 <- validation_set$arr_delay - ridge_5$predict(validation_set[index])
RMSE_5 <- sqrt(mean(resid_5^2)) 

resid_10 <- validation_set$arr_delay - ridge_10$predict(validation_set[index])
RMSE_10 <- sqrt(mean(resid_10^2)) 

resid_50 <- validation_set$arr_delay - ridge_50$predict(validation_set[index])
RMSE_50 <- sqrt(mean(resid_50^2)) 

resid_200 <- validation_set$arr_delay - ridge_200$predict(validation_set[index])
RMSE_200 <- sqrt(mean(resid_200^2)) 

resid_10000 <- validation_set$arr_delay - ridge_10000$predict(validation_set[index])
RMSE_10000 <- sqrt(mean(resid_10000^2)) 

RMSE <- c(RMSE_1, RMSE_5, RMSE_10, RMSE_50, RMSE_200, RMSE_10000)
RMSE


```

We found that the optimal value for λ is 200.


## 5

We predict the test set here.

```{r}
ridge_mod <- ridgereg(formula, test_set, lambda=200)
resid_mod <- test_set$arr_delay - ridge_mod$predict(test_set[index])
RMSE_mod <- sqrt(mean(resid_mod^2)) 

RMSE_mod  
```


The RMSE of ours predicted model is 17.99421.

